\documentclass[11pt]{psuthesis}


\usepackage{amsmath,amssymb}
\usepackage[square]{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{longtable}
\usepackage{pdflscape}

%% Fix the text citations so that there is no comma between the authors and 
%% year.  This will help contain the furious Brandt red pen.
%\bibpunct{(}{)}{;}{a}{}{,}
\bibpunct{[}{]}{;}{a}{,}{,}

\renewcommand{\thesubsubsection}{(\roman{subsubsection})}

% usage - \mark{myfootnote}{This is my footnote} and 
% \recall{myfootnote}

%this defines where your images for the build are stored.
%\graphicspath{{images/}}

%% Change the fonts back to something reasonable.
%% Note: scriptsize is typically smaller than footnotesize.
%\renewcommand{\scriptsize{\@setfontsize\scriptsize\@ixpt{9pt}}
%\renewcommand{\footnotesize{\@setfontsize\scriptsize\@xpt{10pt}}

%\thesisdraft
%% Uncomment the above line to generate a copy of the thesis that has the phrase:
%%"Draft: <current date> at <current time>" to be printed in the head of each page.

%% This is just to set your degree. Set the toggle within the document (below).
\newtoggle{masters}

\begin{document}

% Title information
\author{John R. Leeman}

\title{\uppercase {Laboratory Studies of Fault Stability and Slow Earthquakes}}
\dept{College of Earth and Mineral Sciences}
\major{Geosciences}

\togglefalse{masters}

\submitdate{May 2017}

\copyrightyear{2017}

%
%\begin{singlespace}

\readerone{Chris Marone \\
         \prof{Geosciences} \\
         \adviser \\
         \chair
}

\readertwo{Demian Saffer\\
           \adviser \\
           \prof{Geosciences}
}

\readerthree{Sridhar Anandakrishanan\\
             \prof{Geosciences}
}

\readerfour{Richard Alley  \\
	  \prof{Geosciences}
} 

\readerfive{Derek Elsworth  \\
	  \prof{Energy and Mineral Engineering}
} 

\readersix{Lee Kump  \\
	  \head{Geosciences}
} 
%%   Key to titles:
%% Associate Professor: \asocprof{of what}
%% Assistant Professor: \assistprof{of what}
%% Full Professor: \prof{of what}
%% also Dept. Head: \head{of what}
%% You can also do things like: ``Associate \head{of what}''

\begin{frontmatter}

%this is the ``normal'' signature page from the original version of the class - the official copy of the thesis or dissertation does not contain signatures of committee members, so omit the signature page. The committee page is a necessary evil.
%\signaturepage

\begin{doublespace}
\titlepage
\end{doublespace}

%this is the new committee page
\committeepage

\abstract

Fault zones are areas of localized deformation that accommodate strain in the Earth's crust accumulated over time due to tectonic motion or stress transfer from adjacent areas. Faults are traditionally considered to accommodate this strain by either constant slow movement (creep), or by rapid catastrophic failure events (earthquakes). The behavior of faults which produce earthquakes has been extensively studied, including characterization of the time and slip predictability of earthquakes, frequency-magnitude distributions, aftershock decay patterns, dynamic triggering, and frictional processes. Chance observations in global positioning system (GPS) data from the Vancouver Island area in 2001 revealed a new kind of fault slip that had not been considered before, slow-slip events. Since those early observations, slow-slip events have been observed at most major subduction interfaces and even in glacial systems. 

In this dissertation, I strive to answer some of the fundamental questions about slow-slip systems. Little is known about the dynamics of these systems and how they operate. Scattered laboratory observations have provided clues, but this study is the first systematic examination of slow-slip earthquakes and their frictional behavior in the laboratory. I examine questions such as what controls how a fault zone will fail, what the velocity and normal stress sensitives are, and connect those mechanisms to observations from a natural slow-slip system beneath Whillans Ice Stream in western Antarctica.

In chapter 1, I demonstrate how to modify the stiffness of the testing machine to create slow-slip events in the laboratory in an artificial granular material. I also present a method to automatically calculate the stiffness of each slip event in a given experiment. Chapter 2 extends this work into a synthetic fault gouge material and carefully examines the sensitivity of the system to the stiffness of the testing apparatus. Chapter 3 introduces the new parameter of velocity dependence into the test suite and demonstrates that designer frictional laws (those whose velocity dependence of friction is itself velocity dependent) are not necessary to explain observations of slow-slip. Chapter 4 introduces the field area of Whillans Ice Stream, a system that hosts slow-slip events daily at an ice-till interface approximately 1 km below the ice surface. I test samples of the till obtained by piston coring of the ice stream in 1989 and develop a simple hydrologic model to determine the potential stress states of the system and an effective medium model to predict the acoustic velocities under those stresses. The entire system is then examined in the light of stability theory to postulate why it has been able to remain in the slow-slip regime for as long as it has been observed. Finally, in chapter 5 I examine the generation of electrical potential differences on slipping experimental faults and critically evaluate the generation mechanisms proposed in the literature.

This dissertation provides insight into the mechanisms and controls of fault slip. I demonstrate that fault failure is not a bifurcation between stable and unstable, but rather a continuous spectrum of failure modes from slow to fast stick-slip. The evidence provided shows that the stiffness of the system is the dominant controlling mechanism and that higher order frictional terms are not required to explain the basic spectrum of behaviors observed.

%% this is the end of the abstract

%% after the abstract come the table of contents, the list of tables,
%% and the list of figures
%% Note about the figure list... so the figure list is a decent length,
%% use the following command for the figure captions:
%% \caption[Short figure title to appear in figure list]{Normal figure caption}
%% (this trick unfortunately does not work with the deluxetable 
%% ``\tablecaption'' command, so be careful what you put in the table captions)


\tables


\acknowledgments  

I am forever indebted to my parents, Russell and Dianne Leeman for their lifelong support of my curiosity and for teaching me the value of hard work. Doctoral research is impossible without these. My wife, Lendi, has been more supportive and understanding that I ever knew was possible. 

My colleagues in the department of geoscience provided many stimulating discussions. I especially would like to thank Dr. Charles Ammon for the large amounts of time we spent in discussion on many topics and side projects. Our talks broadened my interests and led me to a deeper understanding and appreciation of many subjects.

Last, but not least, thank you to all members of my committee. Without your guidance, help, and advice the work presented here would not have been possible. All of the work presented here is cross-disciplinary because of your collaboration, patience, and willingness to embark on new lines of inquiry.

%\preface 

\clearpage

\vspace*{2.0truein}

\parbox{4.0truein}{
\par\noindent
Somewhere, something incredible is waiting to be known.\\
\hspace*{\fill}--Carl Sagan
}

\vspace{4pc}

\parbox{4.0truein}{
\par\noindent
Out of the cradle onto the dry land here it is standing... atoms with consciousness... matter with curiosity.\\
Stands at the sea... wonders at wondering... I... a universe of atoms... an atom in the universe.\\
\hspace*{\fill}--Richard P. Feynman
}
%\normalsize
\end{frontmatter}

%% this is the end of the front matter

%% Chapters
\addcontentsline{toc}{chapter}{Introduction}  
\include{intro}
\include{chap_granular_stiffness/Chap_granular_stiffness}
\include{chap_lab_slow_eq/Chap_lab_slow_eq}
\include{chap_slow_slip_details/Chap_slow_slip_details}
\include{chap_whillans/Chap_whillans}
\include{chap_electrical/Chap_electrical}

%\begin{singlespace}
%\include{conclusions}

%% now we include the appendices
\appendices
\include{appendix_biax_experiments/appendix_biax_experiments}
\include{appendix_triax_experiments/appendix_triax_experiments}
\include{appendix_sample_prep/appendix_sample_prep}
\include{appendix_load_response/appendix_load_response}
\include{appendix_load_calibration/appendix_load_calibration}
\include{appendix_dcdt_calibration/appendix_dcdt_calibration}
\include{appendix_look_format/appendix_look_format}

%% finally comes the bibliography and vita
%% the bibliography is generated automatically using BibTeX
%% Note: in order to get the (author year) citation style, this example 
%% includes a different *.bst style file than psuthesis.bst.  Sarah found 
%% this style file on: http://www.ee.oulu.fi/~harza/latex/  
%% This style includes the paper titles in the bibliography, so use it 
%% if you want:
%\bibliographystyle{/enter/path/to/ayphdthesis_mod}
%% However, according to the grad school thesis guide (c.2004), there is no 
%% special formating required for the bibliography, and to just follow 
%% the citation styles of your field, in that case the ``apj'' style is 
%% prefectly OK, so that is the one that is going to be included in this
%% example file (use whichever style you prefer):
\begin{singlespace}
\bibliographystyle{apa}
\end{singlespace}
%% the following is the path to you *.bib file 
%% (you do not need to enter the ``.bib'' extention)
\bibliography{leeman_refs}

%% the thesis must end with a Curriculum Vitae (**one page or less**)
%% (this is Sarah's formatting, not sure how it compares to the other examples)
\iftoggle{masters}{\clearpage}{
\clearpage
\begin{singlespace}
\vita
\Large
\vspace*{-0.4truein}
\centerline{{\bf John R. Leeman}}

\medskip

\large
\centerline{{\bf Education}}
\normalsize

\smallskip

\par\noindent
\textbf{\textit{The Pennsylvania State University}}\, State College, Pennsylvania\hfill 2012-2017

\smallskip

\par\noindent
\hspace{0.10truein}  
\parbox{6.15truein}{
\par\noindent
Ph.D in Geosciences, May 2017\\  Advisors: Dr. Chris Marone and Dr. Demian Saffer}

\medskip

\par\noindent
\textbf{\textit{University of Oklahoma}}\, Norman, Oklahoma\hfill 2007-2012

\smallskip

\par\noindent
\hspace{0.10truein}  
\parbox{6.15truein}{
\par\noindent
B.S. in Meteorology, B.S. in Geophysics, minor in Mathematics
}

%% Core Competencies
\medskip

\large
\centerline{{\bf Core Competencies}}
\normalsize

\vspace{-0.6cm}
\begin{center}
    \begin{tabular}{l l l}
   Experimental Design & Biaxial \& Triaxial Deformation experiments & Python Programming\\
   Electronics Design & 3D CAD Design & 3D Printing\\
   Public Speaking & Microcontroller Programming & Data Processing\\
   Scientific Visualization & System Administration & Mathematical Modeling\\
    \end{tabular}
\end{center}

%% Selected Grants & Awards
\smallskip

\large
\vspace{-0.4cm}
\centerline{{\bf Selected Grants \& Awards}}
\normalsize

\medskip

\par\noindent
Peter Deines Lectureship \hfill 2016\\
AGU Outstanding Student Paper \hfill 2011, 2015\\
National Science Foundation Graduate Fellow \hfill 2013--2017\\
Shell Geosciences Energy Research Facilities Award \hfill 2013--2016\\
Hess Corporation Exploration \& Production Technology Scholarship \hfill 2014, 2015\\
\medskip

\large
\vspace{-0.6cm}
\centerline{{\bf Selected Publications}}
\normalsize

\noindent\hangindent=1cm JR Leeman, RD Valdez, RB Alley, S Anandakrishnan, and DM Saffer. Mechanical and hydrologic proper- ties of whillans ice stream till: Implications for basal strength and stick-slip failure. Journal of Geophysical Research: Earth Surface, 121(7):1295--1309, 2016.

\noindent\hangindent=1cm JR Leeman, DM Saffer, MM Scuderi, and C Marone. Laboratory observations of slow earthquakes and the spectrum of tectonic fault slip modes. Nature Communications, 7, 2016.

\noindent\hangindent=1cm John Leeman, Marco Maria Scuderi, Chris Marone, and Demian Saffer. Stiffness evolution of granular layers and the origin of repetitive, slow, stick-slip frictional sliding. Granular Matter, pages 1--11, May 2015.

\noindent\hangindent=1cm J. R. Leeman, M. M. Scuderi, C. Marone, D. M. Saffer, and T. Shinbrot. On the origin and evolution of electrical signals during frictional stick slip in sheared granular material. J. Geophys. Res. Solid Earth, 119(5):4253--4268, May 2014.

%\smallskip

%\par\noindent
%\textbf{\textit{Doctoral Research}}\, The Pennsylvania State University\hfill199?--Present
%\par\noindent
%Thesis Advisor: Prof. Someone R. Other

%\smallskip
%
%\par\noindent
%\hspace{0.10truein}  
%\parbox{5.7truein}{
%\par\noindent
%This research involved lots of cool stuff.
%}


\medskip

\large


%
% Professional and Service Activities
%

\centerline{{\bf Professional \& Service Activities}}
\normalsize

\smallskip

\par\noindent
\textbf{\textit{Academic Integrity Committee}}\, The Pennsylvania State University\hfill 2014--2017

\smallskip

\par\noindent
\hspace{0.10truein}  
\parbox{5.7truein}{
\par\noindent
Graduate representative for academic integrity investigations and hearings.
}


\smallskip

\par\noindent
\textbf{\textit{Session Co-Chair}}\, American Geophysical Union\hfill 2014--2016

\smallskip

\par\noindent
\hspace{0.10truein}  
\parbox{5.7truein}{
\par\noindent
Co-chair a session on education and data visualization at the AGU Fall Meeting.
}



\smallskip

\par\noindent
\textbf{\textit{Instructor}}\, The Pennsylvania State University\hfill 2015--2016

\smallskip

\par\noindent
\hspace{0.10truein}  
\parbox{5.7truein}{
\par\noindent
Assist teaching seismology laboratory and develop and teach a 3 hour course on techniques in experimental geoscience.
}



\end{singlespace}
}
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
